TO FIND

### 2013-05-23 ms.tex

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{color,lineno,setspace,graphicx,multirow,kpfonts}
\usepackage[top=2.4cm,left=2.4cm,top=2.4cm,bottom=2.4cm,includefoot]{geometry}
\usepackage[style=bes]{biblatex}
\bibliography{./library}

@@

\textbf{Title:}   Using neutral theory to reveal the contribution of dispersal to community assembly\\

@@

Community assembly is concerned by patterns and processes occurring at various
spatial scales \parencite{Levin1992}. Until the development of metacommunity
ecology, studies on community assembly were often restricted to local
populations, with a strong focus on pairwise interactions (e.g.
\textcite{MacArthur1972, May1973,Pimm1982,DeAngelis1992}). The emphasis on local
communities has been vigorously criticized by \textcite{Ricklefs2008}, who has
long recognized that local dynamics and community structure are strongly
contingent on processes occurring at much larger spatial scales
\parencite{Ricklefs1987}. This perspective is particularly relevant to
limnology, where exchanges of organisms and nutrients affect community and
ecosystem properties from the local (e.g. vertical mixing
\parencite{Ryabov2011}) to the regional (e.g. connection of lakes
\parencite{Leibold2004b} scales. It emphasizes the importance of dispersal
relative to pairwise interactions in the organization of ecological communities.
The metacommunity concept has been proposed by \textcite{Leibold2004a} as a
novel approach to link different spatial scales in ecology. It builds on
feedbacks between local scale processes, such as competitive interactions and
local adaptation, and regional scale processes such as dispersal, gene flow and
speciation. Ecologists are now required to move toward a predictive ecology,
integrating elements of theoretical ecology \parencite{Thuiller2013}, and the
metacommunity perspective appears naturally as the appropriate conceptual
framework to develop the new modeling techniques required to fill this
challenge. The development of neutral theory has been quite provocative in that
respect, as one could see it as a step back in time. Neutral theory makes the
provocative assumption that species are ecologically equivalent and thereby any
variation in the environment has no impact on demography
\parencite{Bell2000,Hubbell2001}. Only demographic stochasticity and dispersal
drive the structure of neutral ecological communities. It therefore appears
that, on first sight, neutral theory is useless.  We will develop in this paper
the argument that neutral theory could be a useful tool to understand the impact
of dispersal on community organization in landscapes of various complexities.

Neutral theory sparked an historical debate in community ecology that is still
lasting after more than a decade \parencite{Chave2004, Etiennee2011,
Rosindell2012,Clark2012}. It was stimulated by the impressive ability of neutral
models to fit several well studied empirical observations such as species
abundance distributions and distance-decay relationships. A remarkable strenght
of neutral theory is to provide a \emph{"formal general theory of abundance
and diversity that will account, in a simple and economical fashion, for the
many patterns that ecologists have documented"} \parencite{Bell2001}. Even if
new studies rejecting neutral theory are consistently published (e.g.
\textcite{Ricklefs2012}), there is now almost a consensus that neutral theory is
a well-developed null hypothesis for niche theory and could even be used as an
adequate approximation of ecological dynamics in some situations.
\textcite{Bell2001} nicely envisionned two perspectives to neutral theory that
are still standing today. Under the weak perspective, neutral theory provides a
set of realistic predictions of community organization despite false
assumptions. Even if being fundamentally wrong, neutral theory is still useful
when used as a null hypothesis \parencite{Gotelli2006}. It is considered as an
improvement over traditional null hypotheses based on randomization
\parencite{Gotelli2000} because it readily integrates dispersal. The strong
version on the other hand posits that neutral theory is a satisfying
approximation to community dynamics and an appropriate theory to explain the
distribution of biodiversity. It implies that the right mechanisms have been
identified and that the consistently observed differences among species do not
impact community organization.

@@

The instrumentalist view of neutral theory raises the question of why it should
be a satisfying approximation despite knowing the pieces are wrong? We see two
potential answers to this question. A first answer might be that stochasticity
of various origins can blur the deterministic differences among species and
promote the ecological drift \parencite{Gravel2011}. Much has been said the
existence of demographic stochasticity, some ecologists even arguing that
neutral models impede progress in community ecology by hidding niche differences
\parencite{Clark2012}, and we therefore will keep this discussion for other
papers. The second answer is that dispersal and historical contingencies might
have a much more profound impact on species distribution \parencite{Bahn2007}
and ecological dynamics. The debate over the equivalence assumption and
demographic stochasticity has perhaps overlook the recognition of how much
dispersal influence community assembly.

In this paper, we argue that neutral theory is useful both to understand and
predict the impact of dispersal on community organization. Even
for purely theoretical analyses, we need a benchmark without niche differences to
reveal the role of dispersal in structuring communities, and understand how it
interacts with niche differentiation. We will explore recent applications of
neutral theory, at the crossroad of network theory, to better represent the
impact of landscape structure on biodiversitity distribution. This analysis will
prove particularly relevant to limnology, where most riverine and lacustre
habitats are characterized by a their discrete nature and spatially complex
arrangements. We will also reveal the relative contribution of ecological
interaction and niche differentiation by contrasting predictions of a neutral
model to other metacommunity perspectives.

Our main objective in this paper is to use neutral theory to stress the
importance of landscape network structure on the distribution of diversity. We
refer to the landscape organization as a "spatial contingency"
\parencite{Peres-Neto2013} that could potentially affect the coexistence
mechanisms at play. We will therefore move from a perspective where dispersal is
either global or spatially explicity (e.g. over a lattice), and spatial
constant, to a perspective focusing on the variance of dispersal. A second
generation of neutral models (e.g. \textcite{Economo2008,
Economo2011,Desjardins2012a,Desjardins2012b}, and even experiments
\parencite{Altermat2012}, recently introduced more realistic lanscapes and found
surprising contributions of spatial contingencies. We will start with a short
review of the main approaches to describe spatial networks and the studies
investigating them. Then we will describe three simple toy models of
metacommunity dynamics, taking this opportunity to review their assumptions and
main predictions. We provide as Supplementary Material the R scripts for the toy
models and all simulations conducted for this paper. We then conduct simple
simulations of these models to reveal the impact of spatial network structure on
diversity distribution. We conclude with a discussion on the operationally of
the framework.

@@

For simulations, spatial networks can easily be generated with the random
geometric graph algorithm \cite{sed01}. In this algorithm, all vertices are
assigned to a position in some two-dimensional space, most often the unit
square. Then, all pairs of vertices within some threshold Euclidean distance $r$
are connected with an edge. The resulting networks have the desirable
property of locality: if a vertex $A$ is connected to two vertices $B$ and $C$.
then $B$ and $C$ are more likely to be connected than two random vertices.
Random geometric networks have been extensively studied \cite{app97a,app97
b,app02a,app02b,pen03} and we provide a R function to generate them. We also
provide the code for another structure that we call the geometric tree of
doom. It builds a tree from the shortest path tree \cite{dij59} of a vertex in
a random geometric network. This random geometric tree of doom looks like a
lake connected by rivers to a series of smaller lakes.

@@

@@

In this section we describe three toy models representing different
perspectives of metacommunity ecology: patch dynamics, neutral dynamics and
species sorting. While the neutral model is interesting in itself, it is by
comparison with a model without any interactions (patch dynamics) and with niche
differentiation (species sorting) that we will be able to fully understand the
interaction between these processes and landscape structure. Despite neutral,
competitive interactions in neutral models are very strong because of the
zero-sum assumption (the community is always at carrying capacity). We will
first review the fundamental assumptions of each model with their description
(Table 1 summarizes the parameters and variables that are used), and then
briefly discuss their main predictions. Simulation results are presented in the
next section, with the corresponding R code provided in the Supplementary
Material.

@@

Predictions of the patch dynamics metacommunity model are quite straightforward.
First, a fundamental result of metapopulation ecology is that persistence will
occur if colonization probability is larger than extinction probability ($c>e$).
Given that all species are the same, then we should expect the regional
diversity to be $S$ if this condition is satisfied and $0$ if not. The situation
is however more complex in spatially explicit landscapes with complex
connectivity matrices \parencite{Hanski1998}. Spatially explicit dispersal
usually reduces the occupancy and thereby the likelihood of persistence. The
second prediction is that, given spatial variation in connectivity, there will
be spatial variation in occurrence probability. Given the above formulation of a
colonization event to occur, the probability that an empty location is colonized
is $I_ix=1-\prod d_x(1-C_{ixy})$. This equation basically tells us that the
colonization probability will increase asymptotically with the degree of a patch
(because of the product). It is easy to show from metapopulation theory that the
occurrence probability in a patch is then $p_ix=I_x(I_ix+e)^{-1}$. The feedback
between local and regional dynamics arises because all $p_ix$ from the landscape
are dependent from each other. Simulation are usually conducted to solve the
model for a large landscape, but numerical solutions are theoretically possible.
The aggregation across the $S$ species of the regional species pool is obtained
by taking the summation of occurrence probabilities over all species, $s_x =
\sum{p_i}$. Because in this model all species are equal, we expect the local
species richness to be a linear function of the patch degree. Multi-species
analysis of metapopulation models also reveals interesting predictions on other
aspects of community organization at various spatial scales such as species-area
relationship \parencite{Hanski1997}, and proved to be useful in conservation
ecology with predictions of extinctions following habitat destruction
\parencite{Nee1994,Rybicki2013}.

@@

Neutral theory introduces strong competitive interactions by assuming there is a
finite number of individuals that could occupy a  patch. There are different
ways to simulate this \emph{zero-sum rule} \parencite{Bell2000,Hubbell2001}, but
they all result in the same constraint that the increase in abundance of a
species could only occur after an equivalent decrease by another species. One
important change in the formulation of most neutral models relative the patch
dynamics model presented above is therefore that it is individual-based, not
population based. We therefore considered in our toy model of neutral dynamics
that each local patch is made of $J_x$ individuals. The model tracts the local
abundance of all species $N_{ix}$ in each local patch. At each time step an
individidual dies with probability $k$. Recruitment only occurs in vacant sites,
similarly to a tree by tree replacement process in a closed canopy forest.

The formulation of the recruitment probability is the central piece of all
neutral models, making possible the coupling with the metacommunity and
neighbouring patches. We adopt a simple formulation in our model based on
\parencite{Gravel2006}. The approach is conceptually similar to placing a seed
trap in a gap in a forest and picking a seed at random among the one that fall
in to determine the identity of the recruited species. The composition of the
seed pool in that trap will be a mixture of local dispersal and immigrants from
the metacommunity. For simplicity, we consider three spatial scales of dispersal
but it would be easy to generalize the approach to a continuous seed dispersal
kernel \parencite{Gravel2006}. The parameter $m$ is the probability that the
recruit is a migrant from neighbouring patches, $M$ is the probability it comes
from a larger (and fixed) metacommunity, and consequently by substraction
$1-m-M$ is the probability it comes from local dispersal. The fraction
$N_{ix}J_x^{-1}$ is the local relative abundance and $P_ix$ is the relative
abundance of species $i$ in the seed pool coming from neigbouring patches $x$.
The relative abundance in the neighborhood is weighted by the degree of the
connected nodes because some nodes will spread their seeds across a higher
number of nodes and thus contribute less to the seed pool. We consider simply
$P_{ix} =\frac{\sum d_y^{-1}P_{iy}}{\sum dy^{-1}}$.  We assume for simplicity
(and without loss of generality, \parencite{Bell2000}) that the relative
abundance in the metacommunity is uniform, ie equal to $S^{-1}$. This
immigration prevents the collapse of the metacommunity because otherwise all
species except one will face extinction by ecological drift (speciation prevents
this phenomenon to occur in \textcite{Hubbell2001}). The local recruitment
probability is consequently $Pr_{ix} = MS^{-1} + mP_{ix} +
(1-m-M)N_{ix}J_x^{-1}$.

The model is neutral because it assumes that the probabilities of a local
recruitment, an immigration and a mortality event are all equal. Demographic
stochasticity is the source of variations in abundance but larger disturbances
could be simulated as well, as long as they hit all species with the same
probability, independently of their density. The fundamental feature of neutral
dynamics is therefore the ecological drift, defined as population changes
emerging from neutraly stable population dynamics. It can be measured as the
variance between replicated time series of community dynamics
\parencite{Gravel2011}. \textcite{Hubbell2001} provides a very comprehensive
analysis of the model, with specific attention to the effect of the different
parameters on the drift (and consequently variance in abundance) and time to
extinction. Despite its simplicity, the neutral model is surprisingly rich in
the predictions it makes. \textcite{Bell2001} and \textcite{Hubbell2001}
analyzed the ability of neutral models to predict species abundance
distributions, range-abundance relationship, spatial variation in abundance,
species-area relationship, community turnover (beta-diversity) and
co-occurrence. Other than the ecological equivalence assumption, one of the most
criticized aspect of neutral models is the realism of the speciation process and
the required speciation rates to sustain species richness
\parencite{Ricklefs2003,Etienne2007}. Recent neutral models with more credible
speciation models \parencite{Rosindell2009,Desjardins2012a} revealed the
difficulty to maintain diversity in neutral models over macro-evolutionary time
scales. These models however also generated interesting novel predictions on
endemic species richness \parencite{Rosindell2011,Desjardins2012b}.

@@

The species-sorting and the mass effect perspectives build on the notion of
species-specific responses to a spatially varying environment
\parencite{Leibold2004a}. There are various ways to simulate such dynamics and
we picked the lottery models, in line with tradition \parencite{Mouquet2002} and
for its proximity to the neutral model described above. Competition for space
occurs during recruitment after the death of an adult. The recruitment is a
lottery against potential candidates. The recruitment probability is however
biased by species specific response to local environmental conditions.

The lottery dynamics described above for the neutral model assume there is a
very large number of offsprings that are candidate for recruitment but only one
will survive and develop to the adult stage. The effect of a differentiation to
local environmental conditions could be implemented at this stage.The $J_x$
individuals all experience unique environmental conditions $E_{nx}$ called a
microsite $n$. We considered a patch average $\overline{E_x}$, with a
within-patch variance $\sigma_x$. The regional average is $\overline{E_R}$ and
the regional variance $\sigma_R$ (for simplicity we considered normal
distributions, but different distributions will lead to different regional
similarity constraints \parencite{Mouquet2003,Tilman2004,Gravel2006}). We
consider that a fraction $\lambda_{inx}$ of offsprings reaching the microsite
where recruitment occurs will survive. The recruitment probability is therefore
biased because only some species will be able to cope with the local
environmental conditions. For tractability we define the relative abundance in
the seed rain as $Z_{ix} = MS^{-1} + mP_{ix} + (1-m-M)N_{ix}J_x^{-1}$. The
calculation of the relative abundance in the seed rain is the same as the
neutral model but the recruitment probability differs because only a fraction of
offspring survive. It is formulated as $Pr_{ix} =
\frac{\lambda_{inx}Z_{ix}}{\sum \lambda_{jnx}Z_{jx}}$. The function describing
the relationship between microsite conditions and survival could take various
forms and we used the traditional gaussian curve describing the niche,
$\lambda_{inx} =  \exp{-\frac{(E_{nx}-u_i)^2}{2\Pi b_i^2}}$, where $u_i$ is the
niche optimum and $b_i$ is niche breadth. Note that the model will converge to a
neutral model when the niche breadth tends to infinity (which is in fact how we
simulated neutral dynamics in the Supplementary Material to minimize the
complexity of the code).

Analyses of similar models with a combination of dispersal and species-sorting
shown that predictions are extremely variables and depend on the distribution of
environmental conditions, niche optimums and breadth. For instance, a
well-studied prediction of neutral models is the species abundance distribution.
It was shown that niche models can predict similar distributions
\parencite{Tilman2004,Gravel2006}. The main prediction is nonetheless that
stable and predictable (meaning which species will coexist) if species are
sufficiently dissimilar. Local species richness will first depend on the joint
effects of local heterogeneity and niche breadth because coexistence requires a
limiting similarity \parencite{Schwilk2005}. Local species richness could be
increased by a mass effect when dispersal is consistently supplying individuals
coming from more favorable locations (refuges). The limiting similarity required
to maintain regional coexistence depends on the amount of dispersal because
exchanges among communities homogenizes environmental conditions. This is one of
the main result from the species sorting theory and a clever example of
local-regional feedbacks: increasing dispersal promotes local coexistence, but
on the other hand it diminishes regional coexistence. Only the best average
competitors will remain at very high dispersal.  We therefore expect a
hump-shaped relationship between dispersal and alpha diversity, with a peak at
intermediate dispersal. On the other hand, we expect a monotonic decrease of
beta and gamma diversity with dispersal \parencite{Mouquet2003}. This prediction
has been validated in some experiments \parencite{Venail2008, Logue2011}.

@@

In fig. 1, we illustrate the outcome of simulations using the neutral model on
the four different network shapes considered in this article. All of these
networks have the same number of nodes (e.g. spatial sampling sites), but both
different number of edges (e.g. dispersal routes between sampling sites) and
patterns of connectivity between nodes. We ask how these differences in
connectivity will shape the emerging properties of the community under the
scenarios represented by each meta-community model.

In fig. 2, we present the richness of each node of the network, as a function of
the centrality of the node, under different assumptions of meta-community
dynamics and network structure. It appears that both in the goeographical and
tree graph, the path dynamics model allows more diversity to persist overall.
However, the local diversity increases with the node centrality, whether it is
measured by the degree or the eigencentrality. All situations examined give a
remarkably congruent signal. The patch dynamics model, \emph{i.e.} when there
are no interactions between species, allows more $\alpha$ diversity. The neutral
model allowed the maintenance of slightly more diversity than the
species-sorting model. In all cases however, it is clear that the richness
maintained tend to increase with the node centrality.

>>>>

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{color,lineno,setspace,multirow,kpfonts}
\usepackage{graphicx}
\usepackage[top=2.4cm,left=2.4cm,top=2.4cm,bottom=2.4cm,includefoot]{geometry}
\usepackage[style=bes]{biblatex}
\bibliography{./library}

% TODO:
% TP, PDP & DG: Complete the discussion
% PDP: Additional paragraph on the metrics
% DG: Figure legends
% DG: Abstract
% DG: Compile references
% TP & PDP: fix bugs I generated!!!

@@

\textbf{Title:}   Using neutral theory to reveal the contribution of dispersal to community assembly in complex landscales\\

@@

Ecologists are required to move toward a predictive ecology,
integrating elements of theoretical ecology \parencite{Thuiller2013}, and the
metacommunity perspective appears naturally as the appropriate conceptual
framework to develop the new modeling techniques required to fill this
challenge. The metacommunity concept builds on 
feedbacks between local scale processes, such as competitive interactions and
local adaptation, and regional scale processes such as dispersal, gene flow and
speciation \textcite{Leibold2004a}. This perspective is particularly relevant to
limnology, where exchanges of organisms and nutrients affect community and
ecosystem properties from the local (e.g. vertical mixing
\parencite{Ryabov2011}) to the regional (e.g. connection of lakes
\parencite{Leibold2004b} scales. It emphasizes the importance of dispersal
relative to pairwise interactions in the organization of ecological communities.

The development of neutral theory has been quite provocative in that
respect, as one could see it as a step back in time. Neutral theory makes the
provocative assumption that species are ecologically equivalent and thereby any
variation in the environment has no impact on demography
\parencite{Bell2000,Hubbell2001}. Only demographic stochasticity and dispersal
drive the structure of neutral ecological communities. It therefore appears
that, on first sight, neutral theory is useless.  We will develop in this paper
the argument that neutral theory could be a useful tool to understand the impact
of dispersal on community organization in landscapes of various complexities.

Neutral theory sparked an historical debate that is still
lasting after more than a decade \parencite{Chave2004, Etiennee2011,
Rosindell2012,Clark2012}. It was stimulated by the surprising ability of neutral
models to fit several well studied empirical observations such as species
abundance distributions and distance-decay relationships. A remarkable strenght
of neutral theory is to provide a \emph{"formal general theory of abundance
and diversity that will account, in a simple and economical fashion, for the
many patterns that ecologists have documented"} \parencite{Bell2001}. Even if
new studies rejecting neutral theory are consistently published (e.g.
\textcite{Ricklefs2012}), there is now almost a consensus that neutral theory is
a well-developed null hypothesis for niche theory and could even be used as an
adequate approximation of ecological dynamics in some situations.
\textcite{Bell2001} nicely envisionned two perspectives to neutral theory that
are still standing today. Under the weak perspective, neutral theory provides a
set of realistic predictions of community organization despite false
assumptions. Even if being fundamentally wrong, neutral theory is still useful
when used as a null hypothesis \parencite{Gotelli2006}. It is considered as an
improvement over traditional null hypotheses based on randomization
\parencite{Gotelli2000} because it readily integrates dispersal. The strong
version on the other hand posits that neutral theory is a satisfying
approximation to community dynamics and an appropriate theory to explain the
distribution of biodiversity. It implies that the right mechanisms have been
identified and that the consistently observed differences among species do not
impact community organization.

@@

The instrumentalist view of neutral theory raises the question of why it should
be a satisfying approximation despite knowing the pieces are wrong? We see two
potential answers to this question. A first answer might be that stochasticity
of various origins can blur the deterministic differences among species and
promote ecological drift \parencite{Gravel2011}. Much has been said the
existence of demographic stochasticity, some ecologists even arguing that
neutral models impede progress in community ecology by hidding niche differences
\parencite{Clark2012}, and we therefore will let this discussion for other
papers. The second answer is that dispersal and historical contingencies might
have a much more profound impact on species distribution \parencite{Bahn2007, Boulangeat2012}
and ecological dynamics. The debate over the equivalence assumption and
demographic stochasticity has perhaps overlook the recognition of how much
dispersal influence community assembly.

In this paper, we argue that neutral theory is useful both to understand and
predict the impact of dispersal on community organization. Even
for purely theoretical analyses, we need a benchmark without niche differences to
reveal the role of dispersal in structuring communities, and understand how it
interacts with niche differentiation. We will explore recent applications of
neutral theory, at the crossroad of network theory, to better represent the
impact of landscape structure on biodiversitity distribution. This analysis will
prove particularly relevant to limnology, where most riverine and lacustre
habitats are characterized by a their discrete nature and spatially complex
arrangements \parencite{Peterson2013}. We will also reveal the relative contribution of ecological
interaction and niche differentiation by contrasting predictions of a neutral
model to other metacommunity perspectives.

Our main objective in this paper is to use neutral theory to stress the
importance of landscape network structure on the distribution of diversity. We
refer to the landscape organization as a "spatial contingency"
\parencite{Peres-Neto2013} that could potentially affect the coexistence
mechanisms at play. We will therefore move from a perspective where dispersal is
either global or spatially explicity (e.g. over a lattice), and spatial
constant, to a perspective focusing on the variance of dispersal. A second
generation of neutral models (e.g. \textcite{Economo2008,
Economo2011,Desjardins2012a,Desjardins2012b}, and even experiments
\parencite{Altermat2012}, recently introduced more realistic lanscapes and found
surprising contributions of spatial contingencies. We will start with a short
review of the main approaches to describe spatial networks and the studies
investigating them. Then we will describe three simple toy models of
metacommunity dynamics, taking this opportunity to review their assumptions and
main predictions. We provide as Supplementary Material the R scripts for the toy
models and all simulations conducted for this paper. We then conduct simple
simulations of these models to reveal with simple examples the impact of spatial network structure on
diversity distribution. We conclude with a discussion on the operationally of
the framework.

@@

For simulations, spatial networks can easily be generated with the random
geometric graph algorithm \cite{sed01}. In this algorithm, all vertices are
assigned to a position in some two-dimensional space, most often the unit
square. Then, all pairs of vertices within some threshold Euclidean distance $r$
are connected with an edge. The resulting networks have the desirable
property of locality: if a vertex $A$ is connected to two vertices $B$ and $C$.
then $B$ and $C$ are more likely to be connected than two random vertices.
Random geometric networks have been extensively studied \cite{app97a,app97
b,app02a,app02b,pen03} and we provide a R function to generate them. The 
position of vertices is typically random, but we could also imagine alterations
where they are either more aggregated or segregated than expected by chance alone.

We also provide the code for a second structure that we call a random geometric tree. 
It builds a tree from the shortest path tree \cite{dij59} of a vertex in a random geometric 
network. This random geometric tree does not exaxctly represent dendritic landscapes 
but is a convenient model to simulate lake connected by rivers to a series of smaller lakes.

@@

% TO PHIL: IT WOULD BE GREAT IF YOU COULD EXPAND ONE OR TWO PARAGRAPHS ON THE 
% MAIN STATISTICS USED TO DESCRIBE SPATIAL GRAPHS. AT LEAST PRESENT THE MAIN 
% DIFFERENCE BETWEEN NETWORK AND VERTEX LEVEL METRICS, AND THEN REFER TO THE 
% TABLE.

@@

In this section we describe three toy models representing different
perspectives of metacommunity ecology: patch dynamics, neutral dynamics and
species sorting. While the neutral model is interesting in itself, it is by
comparison with a model without any interactions (patch dynamics) and with niche
differentiation (species sorting) that we will be able to fully understand the
interaction between these processes and landscape structure. Despite neutral,
competitive interactions in neutral models are very strong because of the
zero-sum assumption (the community is always at carrying capacity). We will
first review the fundamental assumptions of each model with their description
(Table 2 summarizes the parameters and variables that are used), and then
briefly discuss their main predictions. Simulation results are presented in the
next section, with the corresponding R code provided in the Supplementary
Material.

@@

Predictions of the patch dynamics metacommunity model are quite straightforward.
First, a fundamental result of metapopulation ecology is that persistence will
occur if colonization probability is larger than extinction probability ($c>e$).
Given that all species are the same, then we should expect the regional
diversity to be $S$ if this condition is satisfied and $0$ if not. The situation
is however more complex in spatially explicit landscapes with complex
connectivity matrices \parencite{Hanski1998}. Spatially explicit dispersal
usually reduces the occupancy and thereby the likelihood of persistence. The
second prediction is that, given spatial variation in connectivity, there will
be spatial variation in occurrence probability. Given the above formulation of a
colonization event to occur, the probability that an empty location is colonized
is $I_ix=1-\prod d_x(1-C_{ixy})$. This equation basically tells us that the
colonization probability will increase asymptotically with the degree of a patch
(because of the product). It is easy to show from metapopulation theory that the
occurrence probability in a patch is then $p_ix=I_x(I_ix+e)^{-1}$. The feedback
between local and regional dynamics arises because all $p_ix$ from the landscape
are dependent from each other. Simulations are usually conducted to solve the
model for a large landscape, but numerical solutions are theoretically possible.
The aggregation across the $S$ species of the regional species pool is obtained
by taking the summation of occurrence probabilities over all species, $s_x =
\sum{p_i}$. Because in this model all species are equal, we expect the local
species richness to be a linear function of the patch degree (number of edges). Multi-species
analysis of metapopulation models also reveals interesting predictions on other
aspects of community organization at various spatial scales such as the species-area
relationship \parencite{Hanski1997}, and proved to be useful in conservation
ecology with predictions of extinctions following habitat destruction
\parencite{Nee1994,Rybicki2013}.

@@

Neutral theory introduces strong competitive interactions by assuming there is a
finite number of individuals that could occupy a  patch. There are different
ways to simulate this \emph{zero-sum rule} \parencite{Bell2000,Hubbell2001}, but
they all result in the same constraint that the increase in abundance of a
species could only occur after an equivalent decrease by another species. One
important change in the formulation of most neutral models relative the patch
dynamics model presented above is therefore that it is individual-based, not
population based. We therefore considered in our toy model of neutral dynamics
that each local patch holds $J_x$ individuals. The model tracts the local
abundance of all species $N_{ix}$ in each local patch. At each time step an
individidual dies with probability $k$. Recruitment only occurs in vacant sites,
similarly to a tree by tree replacement process in a closed canopy forest.

The formulation of the recruitment probability is the central piece of all
neutral models, making possible the coupling with the metacommunity and
neighbouring patches. We adopt a simple formulation in our model based on
\parencite{Gravel2006}. The approach is conceptually similar to placing a seed
trap in a canopy gap and picking a seed at random among the ones falling
in to determine the identity of the recruited species. The composition of the
seed pool in that trap will be a mixture of local dispersal and immigrants from
the metacommunity. For simplicity, we consider three spatial scales of dispersal
but it would be easy to generalize the approach to a continuous seed dispersal
kernel \parencite{Gravel2006}. The parameter $m$ is the probability that the
recruit is a migrant from neighbouring patches, $M$ is the probability it comes
from a larger (and fixed) metacommunity, and consequently, by substraction,
$1-m-M$ is the probability it comes from local dispersal. The fraction
$N_{ix}J_x^{-1}$ is the local relative abundance and $P_ix$ is the relative
abundance of species $i$ in the seed pool coming from neigbouring patches $x$.
The relative abundance in the neighborhood is weighted by the degree of the
connected nodes because some nodes will spread their seeds across a higher
number of nodes and thus contribute less to the seed pool. We consider simply
$P_{ix} =\frac{\sum d_y^{-1}P_{iy}}{\sum dy^{-1}}$.  We assume for simplicity
(and without loss of generality, \parencite{Bell2000}) that the relative
abundance in the metacommunity is uniform, i.-e. equal to $S^{-1}$. This
immigration prevents the collapse of the metacommunity because otherwise all
species except one will face extinction by ecological drift (speciation prevents
this phenomenon to occur in \textcite{Hubbell2001}). The local recruitment
probability is consequently $Pr_{ix} = MS^{-1} + mP_{ix} +
(1-m-M)N_{ix}J_x^{-1}$.

The model is neutral because it assumes that the probabilities of a local
recruitment, an immigration and a mortality event are all equal across species. Demographic
stochasticity is the source of variations in abundance, but larger disturbances
could be simulated as well, as long as they hit all species with the same
probability, independently of their density. The fundamental feature of neutral
dynamics is therefore the ecological drift, defined as population changes
emerging from neutraly stable population dynamics. It can be measured as the
variance between replicated time series of community dynamics
\parencite{Gravel2011}. \textcite{Hubbell2001} provides a very comprehensive
analysis of the model, with specific attention to the effect of the different
parameters on the drift (and consequently variance in abundance) and time to
extinction. Despite its simplicity, the neutral model is surprisingly rich in
the predictions it makes. \textcite{Bell2001} and \textcite{Hubbell2001}
analyzed the performance of neutral models to predict species abundance
distributions, range-abundance relationship, spatial variation in abundance,
species-area relationship, community turnover (beta-diversity) and
co-occurrence. Recent trophic neutral models also predicted realistic ecological networks \parenciteCanard2012}. Other than the ecological equivalence assumption, one of the most
criticized aspect of neutral models is the realism of the speciation process and
the required speciation rates to sustain species richness
\parencite{Ricklefs2003,Etienne2007}. Recent neutral models with more credible
speciation models \parencite{Rosindell2009,Desjardins2012a} revealed the
difficulty to maintain diversity in neutral models over macro-evolutionary time
scales. These models however also generated interesting novel predictions on
endemic species richness \parencite{Rosindell2011,Desjardins2012b}.

@@

The species-sorting and the mass effect perspectives build on the notion of
species-specific responses to a spatially varying environment
\parencite{Leibold2004a}. There are various ways to simulate such dynamics and
we picked the lottery model, in line with tradition \parencite{Mouquet2002} and
for its proximity to the neutral model described above. Competition for space
occurs during recruitment after the death of an adult. The recruitment is a
lottery among potential candidates as in the neutral model. The recruitment probability 
is however biased by species specific response to local environmental conditions.

The lottery dynamics described above for the neutral model assume there is a
very large number of offsprings that are candidate for recruitment but only one
will survive and develop to the adult stage. The effect of a differentiation to
local environmental conditions could be implemented at this stage with a biased 
survival probability. The $J_x$ individuals all experience a unique environmental condition 
$E_{nx}$ called a microsite $n$. We considered a patch average $\overline{E_x}$, with a
within-patch variance $\sigma_x$. The regional average is $\overline{E_R}$ and
the regional variance $\sigma_R$ (for simplicity we considered normal
distributions, but different distributions will lead to different regional
similarity constraints \parencite{Mouquet2003,Tilman2004,Gravel2006}). We
consider that a fraction $\lambda_{inx}$ of offsprings reaching the microsite
where recruitment occurs will survive. The recruitment probability is therefore
biased in favour of the species with highest survival because only some species will be able to cope with the local environmental conditions. For tractability we define the relative abundance in
the seed rain as $Z_{ix} = MS^{-1} + mP_{ix} + (1-m-M)N_{ix}J_x^{-1}$. The
calculation of the relative abundance in the seed rain is the same as the
neutral model but the recruitment probability differs because only a fraction of
offspring survive. It is formulated as $Pr_{ix} =
\frac{\lambda_{inx}Z_{ix}}{\sum \lambda_{jnx}Z_{jx}}$. The function describing
the relationship between a microsite condition and survival could take various
forms; we used the traditional gaussian curve describing the niche,
$\lambda_{inx} =  \exp{-\frac{(E_{nx}-u_i)^2}{2\Pi b_i^2}}$, where $u_i$ is the
niche optimum and $b_i$ is niche breadth. Note that the model will converge to a
neutral model when the niche breadth tends to infinity (which is in fact how we
simulated neutral dynamics in the Supplementary Material to minimize the
complexity of the code).

Analyses of similar models with a combination of dispersal and species-sorting
shown that predictions are extremely variables and depend on the distribution of
environmental conditions, niche optimums and breadth. For instance, a
well-studied prediction of neutral models is the species abundance distribution.
It was shown that niche models can predict similar distributions given appropriate parameters
\parencite{Tilman2004,Gravel2006}. The main prediction is nonetheless that
stable and predictable (meaning which species will coexist) if species are
sufficiently dissimilar. Local species richness will first depend on the joint
effects of local heterogeneity and niche breadth because coexistence requires a
sufficient dissimilaritya mong species \parencite{Schwilk2005}. Local species richness could be
increased by a mass effect when dispersal is consistently supplying individuals
coming from more favorable locations (refuges). The limiting similarity required
to maintain regional coexistence depends on the amount of dispersal because
exchanges among communities homogenizes environmental conditions. This is one of
the main result from the species sorting theory and a clever example of
local-regional feedbacks: increasing dispersal promotes local coexistence, but
on the other hand it diminishes regional coexistence. Only the best average
competitors will remain at very high dispersal.  We therefore expect a
hump-shaped relationship between dispersal and alpha diversity, with a peak at
intermediate dispersal. On the other hand, we expect a monotonic decrease of
beta and gamma diversity with dispersal \parencite{Mouquet2003}. This prediction
has been validated in some experiments \parencite{Venail2008, Logue2011}.

@@

In this section we provide simple simulation results to illustrate the impact of saptial contingencies on species distribution and coexistence. We consider four different landscapes, illustrated at Fig. 1.
with the outcome of simulations using the neutral model. All of these
networks have the same number of nodes (e.g. spatial sampling sites), but both
different number of edges (e.g. dispersal routes between sampling sites) and
patterns of connectivity between nodes. We ask how these differences in
connectivity will shape the emerging properties of the community under the
scenarios represented by each meta-community model. Our analysis is not exhaustive, it is provided simply to illustrate the interaction between metacommunity perspectives and landscape structures on alpha, beta and gamma diversity.

In Fig. 2, we present the species richness of each node of the network (alpha diversity), as a function of
the centrality of the node, under different assumptions of metacommunity
dynamics and network structure. We scaled the species richness by the maximal local species richness to 
facilitate comparison between models. The model parameterization is responsible for differences in both alpha and gamma diversity.  It appears that both in the goeographical and
tree graph, the path dynamics model has a much more considerable variation in local species richness. 
However, in all cases the alpha diversity increases with the node centrality, whether it is
measured by the degree or the eigencentrality. All situations examined give a
remarkably congruent signal.

% TO CHANGE WITH NEW FIGURE (SCALED DIVERSITY)
%The patch dynamics model, \emph{i.e.} when there
%are no interactions between species, allows more $\alpha$ diversity. The neutral
%model allowed the maintenance of slightly more diversity than the
%species-sorting model. In all cases however, it is clear that the richness
%maintained tend to increase with the node centrality.
